---
title: "NYC"
format: docx
editor: visual
---

## Libraries

```{r}

#install.packages('tidyverse')
  library(tidyverse)
#install.packages('dplyr')
  library(dplyr)
#install.packages('gmodels')
  library(gmodels)
#install.packages('readxl')
  library(readxl)
#install.packages('lubridate')
  library(lubridate)
#install.packages('tableone')
  library(tableone)
#install.packages('knitr')
  library(knitr)

```

## Dataset

Mutating Date Column

```{r}

data2 <- read.csv("previous_years.csv")

data1 <- read.csv("edheat2025_live.csv")

data <- bind_rows(data2, data1)
#Combines data sets with similar rows and columns

data$Date <- as.Date(data$END_DATE)
#Turns end date column into date

data$END_DATE <- NULL

```

## Create Temperature Ranges

```{r}

data$TEMP_RANGE <- cut(data$MAX_DAILY_TEMP,
                       
                       breaks = c(-Inf, 60, 80, 90, Inf),
                       
                       labels = c("Cold", "Warm", "Hot", "Extreme Heat")
)

```

## Descriptive Statistics

```{r}

# subset(data, as.integer(format(Date, "%Y")) == 2018)

for (i in 2017:2025) {
  print(summary(subset(data, as.integer(format(Date, "%Y")) == i)))
}

ds_table <- CreateTableOne(data = data)
kable(print(ds_table, showAllLevels = TRUE, printToggle = FALSE, noSpaces = TRUE))

print(ds_table)

```

## Figures just to see

```{r}


plot(data$Date, data$MAX_DAILY_TEMP, 
     main = "Max Daily Temp vs. Time",  
     xlab = "Date",     
     ylab = "Max Daily Temp",        
     col = "blue",                
     pch = 19,                     
     type = "b")


 
```

```{r}
doy <- as.numeric(format(data$Date, "%j"))

plot(doy[idx], data$MAX_DAILY_TEMP[idx],
     type = "l",
     col = "blue",
     xlab = "Day of Year",
     ylab = "Max Daily Temp",
     main = "Max Daily Temperature: 2018 vs 2019",
     ylim = range(data$MAX_DAILY_TEMP[c(idx, idx2)], na.rm = TRUE))

lines(doy[idx2], data$MAX_DAILY_TEMP[idx2],
      col = "green")

legend("topright",
       legend = c("2018", "2019"),
       col = c("blue", "green"),
       lty = 1)
```

```{r}
idx <- format(data$Date, "%Y") == "2017"
idx2 <- format(data$Date, "%Y") == "2025"

# Create dates with the same year (2000)
date2018 <- as.Date(paste0("2000-", format(data$Date[idx], "%m-%d")))
date2025 <- as.Date(paste0("2000-", format(data$Date[idx2], "%m-%d")))

plot(date2018, data$MAX_DAILY_TEMP[idx],
     type = "l",
     col = "blue",
     xlab = "Date",
     ylab = "Max Daily Temp",
     main = "Max Daily Temperature: 2017 vs 2025",
     ylim = range(data$MAX_DAILY_TEMP[c(idx, idx2)], na.rm = TRUE))

lines(date2025, data$MAX_DAILY_TEMP[idx2],
      col = "green")

legend("topright",
       legend = c("2017", "2025"),
       col = c("blue", "green"),
       lty = 1)
```

```{r}
# install.packages("ggplot2")
library(ggplot2) 
# Create a sample categorical vector

# Calculate the frequency table
range_counts <- table(data$TEMP_RANGE)

# Create the bar plot
barplot(range_counts, 
        main = "Frequency of Temp Ranges", 
        xlab = "Temp Range in Degrees F", 
        ylab = "Count", 
        names.arg = c("Below 60", "60-79", "80-89", "Equal or above 90"),
        col = c("blue", "lightblue", "orange", "red"))

```

## Scatter plot of temp per year

```{r}
library(ggplot2)
#install.packages('patchwork')
library(patchwork)

idx <- format(data$Date, "%Y") == "2017"
idx2 <- format(data$Date, "%Y") == "2018"
idx3 <- format(data$Date, "%Y") == "2019"
idx4 <- format(data$Date, "%Y") == "2020"
idx5 <- format(data$Date, "%Y") == "2021"
idx6 <- format(data$Date, "%Y") == "2022"
idx7 <- format(data$Date, "%Y") == "2023"
idx8 <- format(data$Date, "%Y") == "2024"
idx9 <- format(data$Date, "%Y") == "2025"

date2017 <- as.Date(paste0("2000-", format(data$Date[idx], "%m-%d")))
date2018 <- as.Date(paste0("2000-", format(data$Date[idx2], "%m-%d")))
date2019 <- as.Date(paste0("2000-", format(data$Date[idx3], "%m-%d")))
date2020 <- as.Date(paste0("2000-", format(data$Date[idx4], "%m-%d")))
date2021 <- as.Date(paste0("2000-", format(data$Date[idx5], "%m-%d")))
date2022 <- as.Date(paste0("2000-", format(data$Date[idx6], "%m-%d")))
date2023 <- as.Date(paste0("2000-", format(data$Date[idx7], "%m-%d")))
date2024 <- as.Date(paste0("2000-", format(data$Date[idx8], "%m-%d")))
date2025 <- as.Date(paste0("2000-", format(data$Date[idx9], "%m-%d")))

df2017 <- data.frame(
  Date = date2017,
  Temp = data$MAX_DAILY_TEMP[idx]
)
df2018 <- data.frame(
  Date = date2018,
  Temp = data$MAX_DAILY_TEMP[idx2]
)
df2019 <- data.frame(
  Date = date2019,
  Temp = data$MAX_DAILY_TEMP[idx3]
)
df2020 <- data.frame(
  Date = date2020,
  Temp = data$MAX_DAILY_TEMP[idx4]
)
df2021 <- data.frame(
  Date = date2021,
  Temp = data$MAX_DAILY_TEMP[idx5]
)
df2022 <- data.frame(
  Date = date2022,
  Temp = data$MAX_DAILY_TEMP[idx6]
)
df2023 <- data.frame(
  Date = date2023,
  Temp = data$MAX_DAILY_TEMP[idx7]
)
df2024 <- data.frame(
  Date = date2024,
  Temp = data$MAX_DAILY_TEMP[idx8]
)
df2025 <- data.frame(
  Date = date2025,
  Temp = data$MAX_DAILY_TEMP[idx9]
)

# Create individual plots
p1 <- ggplot(df2017, aes(x = Date, y = Temp)) +
  geom_line() + labs(x='2017', y = NULL)
p2 <- ggplot(df2018, aes(x = Date, y = Temp)) +
  geom_line() + labs(x='2018', y = NULL)
p3 <- ggplot(df2019, aes(x = Date, y = Temp)) +
  geom_line() + labs(x='2019', y = NULL)
p4 <- ggplot(df2020, aes(x = Date, y = Temp)) +
  geom_line() + labs(x='2020', y = NULL)
p5 <- ggplot(df2021, aes(x = Date, y = Temp)) +
  geom_line() + labs(x='2021', y = NULL)
p6 <- ggplot(df2022, aes(x = Date, y = Temp)) +
  geom_line() + labs(x='2022', y = NULL)
p7 <- ggplot(df2023, aes(x = Date, y = Temp)) +
  geom_line() + labs(x='2023', y = NULL)
p8 <- ggplot(df2024, aes(x = Date, y = Temp)) +
  geom_line() + labs(x='2024', y = NULL)
p9 <- ggplot(df2025, aes(x = Date, y = Temp)) +
  geom_line() + labs(x='2025', y = NULL)
library(grid)

yleft <- wrap_elements(
  textGrob("Temperature (°F)", rot = 90,
           gp = gpar(fontsize = 14))
)

final_plot <- yleft +
  (
    (p1 + p2 + p3) /
      (p4 + p5 + p6) /
      (p7 + p8 + p9)
  ) +
  plot_layout(widths = c(0.05, 1)) +
  plot_annotation(
    title = "Maximum Daily Temperature by Year"
  )

final_plot
```

## ED visits vs. Mean temperature

```{r}
ggplot(data, aes(x = MAX_DAILY_TEMP , y = HEAT_ED_VISIT_COUNT)) +
  # geom_point adds the data markers
  geom_point(color = "#2C3E50", linewidth = 3.5) +
  geom_point(data = subset(data, HEAT_ED_VISIT_COUNT > 100), color = "red", linewidth = 3.5) +
  
  # Clean up visual layout
  labs(
    title = "Emergency Department Visits vs. Max Daily Temp (2017-2025)",
    x = "Maximum Daily Temperature (°F)",
    y = "Emergency Department Visits (countz)"
  ) +
  theme_minimal()
  
```

## Now we sort by month

```{r}
ggplot(data, aes(x = MAX_DAILY_TEMP, y = HEAT_ED_VISIT_COUNT)) +
  geom_point(size = 3.5) +
  
  geom_point(
    data = subset(data, format(Date, "%m") == "05"),
    aes(color = "May"),
    linewidth = 3.5
  ) +
  geom_point(
    data = subset(data, format(Date, "%m") == "06"),
    aes(color = "June"),
    linewidth = 3.5
  ) +
  geom_point(
    data = subset(data, format(Date, "%m") == "07"),
    aes(color = "July"),
    linewidth = 3.5
  ) +
  geom_point(
    data = subset(data, format(Date, "%m") == "08"),
    aes(color = "August"),
    linewidth = 3.5
  ) +
  geom_point(
    data = subset(data, format(Date, "%m") == "09"),
    aes(color = "September"),
    linewidth = 3.5
  ) +
  scale_color_manual(
    name = "Month",
    values = c(
      "May" = "blue",
      "June" = "red",
      "July" = "orange",
      "August" = "yellow",
      "September" = "lightblue"
    )
  ) +
  labs(
    title = "Emergency Department Visits vs. Max Daily Temp (2017–2025)",
    x = "Maximum Daily Temperature (°F)",
    y = "Emergency Department Visits (count)"
  ) +
  theme_minimal()
```

## Get the four outliers

```{r}
data[data$HEAT_ED_VISIT_COUNT > 100, ]

```

## Zero in on July 2019

```{r}
#dx_2019 <- format(data$Date, "%Y-%m") == "2019-07"
# or
idx_2019 <- format(data$Date, "%Y%m") == "201907"

date_2019 <- as.Date(
  paste0("2000-", format(data$Date[idx_2019], "%m-%d"))
)

df_2019 <- data.frame(
  Date = date_2019,
  Temp = data$MAX_DAILY_TEMP[idx_2019]
)
p1 <- ggplot(df_2019, aes(x = Date, y = Temp)) +
  geom_line() + labs(x='July 2019', y = 'Temp')

p1

```

## ED Visits and Temp vs. Time

```{r, fig.width=15, fig.height=6}
# install.packages("hrbrthemes")
coeff <- max(data$HEAT_ED_VISIT_COUNT, na.rm = TRUE) /
  max(data$MAX_DAILY_TEMP, na.rm = TRUE)

# A few constants
temperatureColor <- "#69b3a2"
priceColor <- rgb(0.2, 0.6, 0.9, 1)

ggplot(data, aes(x=Date, group = year(Date))) +
  
  geom_line( aes(y=HEAT_ED_VISIT_COUNT), size=.5, color=temperatureColor) + 
  geom_line( aes(y=MAX_DAILY_TEMP*coeff), size=.5, color=priceColor) +
  
  scale_y_continuous(
    
    # Features of the first axis
    name = "Emergency Department Visits (countz)",
    
    # Add a second axis and specify its features
    sec.axis = sec_axis(~./coeff, name="Maximum Daily Temperature (°F)")
  ) + 
  
  theme_ipsum() +
  
  theme(
    axis.title.y = element_text(color = temperatureColor, size=13),
    axis.title.y.right = element_text(color = priceColor, size=13)
  ) +
  
  ggtitle("ED Visits and Max Daily Temp vs. Time")
```

## Linear Regression??

```{r}
model.linear <-  lm(HEAT_ED_VISIT_COUNT ~ MAX_DAILY_TEMP, data = data)
summary(model)
```

```{r}
model.pois <- glm(
  HEAT_ED_VISIT_COUNT ~ MAX_DAILY_TEMP,
  family = poisson,
  data = data
)

summary(model.pois)
```

## Plotting linear vs Poisson

```{r}
plot(data$MAX_DAILY_TEMP,
     data$HEAT_ED_VISIT_COUNT,
     pch = 16,
     col = rgb(0,0,0,0.3),
     xlab = "Maximum Daily Temperature (°F)",
     ylab = "Heat ED Visit Count",
     main = "Linear vs Poisson Regression")

abline(model.linear,
       col = "blue",
       lwd = 3)

temp_seq <- seq(min(data$MAX_DAILY_TEMP),
                max(data$MAX_DAILY_TEMP),
                length.out = 100)

pois_pred <- predict(
  model.pois,
  newdata = data.frame(MAX_DAILY_TEMP = temp_seq),
  type = "response"
)

lines(temp_seq,
      pois_pred,
      col = "red",
      lwd = 3)

legend("topleft",
       legend = c("Linear Regression",
                  "Poisson Regression"),
       col = c("blue", "red"),
       lwd = 3)
```

## Pearson and Spearman Correlation

```{r}
# Correlation test between two variables
pearson_corr <- cor.test(data$MAX_DAILY_TEMP, data$HEAT_ED_VISIT_COUNT, method = "pearson")

# View the full results
print(pearson_corr)
```

Wow so this is telling us that the p-value is very small, very close to zero which could be telling us that there is a correlation between temp and ED visits.

```{r}
spearman_corr <- cor.test(data$MAX_DAILY_TEMP, data$HEAT_ED_VISIT_COUNT, method = "spearman", exact = FALSE)
print(spearman_corr)
```
