---
title: "Analysis"
author: "LZM"
date: "2026-06-30"
output: html_document
---


WD
```{r}
getwd()
```

Merging Data - DOESN'T WORK WITH CONFOUNDER CODE - 1

Merge Data
```{r}
library(dplyr)

# Read the data
data2 <- read.csv("previous_years.csv", stringsAsFactors = FALSE)
data1 <- read.csv("edheat2025_live.csv", stringsAsFactors = FALSE)

# Combine the datasets
d <- bind_rows(data2, data1)

# Convert END_DATE to a Date object and name it 'date'
d$date <- as.Date(d$END_DATE)

# Remove the original column if you no longer need it
d$END_DATE <- NULL
```



Step 4: Crude Analyses
Correlation Analysis
To evaluate the crude (unadjusted) relation between temperature and heat-related ED visits, calculate:
Pearson correlation|
```{r}
cor.test(
  d$MAX_DAILY_TEMP,
  d$HEAT_ED_VISIT_COUNT,
  method = "pearson"
)
```


Spearman correlation

```{r}
cor.test(
  d$MAX_DAILY_TEMP,
  d$HEAT_ED_VISIT_COUNT,
  method = "spearman"
)
```



Simple Regression
Fit a crude model: Heat-related ED visits = f(temperature)
You might explore different possible approaches, Linear regression
```{r}
simplelinear <- lm(HEAT_ED_VISIT_COUNT ~ MAX_DAILY_TEMP, data = d)

#Graphing Scatter Plot
plot(
  d$MAX_DAILY_TEMP,
  d$HEAT_ED_VISIT_COUNT,
  main = "Simple Linear Regression",
  xlab = "Maximum Daily Temperature",
  ylab = "Heat ED Visit Count",
  pch = 16,
  col = "black"
)

# Add regression line
abline(simplelinear, col = "red", lwd = 2)
```

Summary L
```{r}
summary(simplelinear)
```


Poisson regression
```{r}
#Poisson regression
pois_model <- glm(
  HEAT_ED_VISIT_COUNT ~ MAX_DAILY_TEMP,
  family = poisson(link = "log"),
  data = d
)

summary(pois_model)

#Graph 
plot(
  d$MAX_DAILY_TEMP,
  d$HEAT_ED_VISIT_COUNT,
  pch = 16,
  col = "black",
  xlab = "Maximum Daily Temperature",
  ylab = "Heat ED Visit Count",
  main = "Poisson Regression"
)

#Line
lines(
  sort(d$MAX_DAILY_TEMP),
  fitted(pois_model)[order(d$MAX_DAILY_TEMP)],
  col = "red",
  lwd = 2
)

```

Summary P
```{r}
summary(pois_model)
```








Quasi Poisson regression
```{r}
#Quasi-Poisson regression
quasi_pois_model <- glm(
  HEAT_ED_VISIT_COUNT ~ MAX_DAILY_TEMP,
  family = quasipoisson(link = "log"),
  data = d
)

summary(quasi_pois_model)

#Graphing 
plot(
  d$MAX_DAILY_TEMP,
  d$HEAT_ED_VISIT_COUNT,
  pch = 16,
  col = "black",
  xlab = "Maximum Daily Temperature",
  ylab = "Heat ED Visit Count",
  main = "Quasi-Poisson Regression"
)

#Line
lines(
  sort(d$MAX_DAILY_TEMP),
  fitted(quasi_pois_model)[order(d$MAX_DAILY_TEMP)],
  col = "red",
  lwd = 2
)
```

Summary QP
```{r}
summary(quasi_pois_model)
```





Negative binomial regression
```{r}
# Negative binomial regression
library(MASS)

nb_model <- glm.nb(
  HEAT_ED_VISIT_COUNT ~ MAX_DAILY_TEMP,
  data = d
)

# Graph
plot(
  d$MAX_DAILY_TEMP,
  d$HEAT_ED_VISIT_COUNT,
  pch = 16,
  col = "black",
  xlab = "Maximum Daily Temperature",
  ylab = "Heat ED Visit Count",
  main = "Negative Binomial Regression"
)

# Line
lines(
  sort(d$MAX_DAILY_TEMP),
  fitted(nb_model)[order(d$MAX_DAILY_TEMP)],
  col = "red",
  lwd = 2
)
```

Summary NB
```{r}
summary(nb_model)
```

Discuss advantages and limitations of each approach.





Step 5: Time-Series Analysis
Build and refine your analytical models. Consider whether adjustment for time-varying factors (e.g., potential confounders), seasonality, temporal trends, or other covariates is warranted. To adjust for time, consider adding variables such as year, day of season, long-term trend. Clearly justify your modeling choices and discuss any assumptions made.


Identified Holidays 
- Independence Day - JULY 4
- Juneteenth - JUNE 19
- Memorial Day - LAST MON. MAY 
- Labor Day - 1ST MON. SEP
- Mother's Day - 2ND SUN. MAY 
- Father's Day - 3RD SUN. JUNE


Example Code 2 - Holidays 
```{r}
# ensure date is in Date format
d$date <- as.Date(d$date)

# create a couple calculated date variables
# year
d$yr <- as.integer(format(d$date, "%Y"))

# month
d$mo <- as.integer(format(d$date, "%m"))

# day of week
d$dow <- weekdays(d$date)
d$dow <- factor(d$dow, levels=unique(d$dow))

# day of the warm season (integer, restarts at 1 each year)
d$season_day <- as.integer(d$date - as.Date(paste0(format(d$date, "%Y"), "-05-01"))) + 1

# create indicator for three summer holidays
# July 4
d$july4 <- ifelse(format(d$date, "%m-%d") == "07-04",1,0)

#ADDING JUNETEENTH
d$juneteenth <- ifelse(format(d$date, "%m-%d") == "06-19",1,0)

# Memorial Day (last Monday in May)
memorial <- as.Date(sapply(d$yr, function(y) {
  x <- as.Date(paste0(y, "-05-31"))
  x - as.POSIXlt(x)$wday   # subtract 0-6 days to reach Monday
}))
d$memorial <- ifelse(d$date == memorial,1,0)

# Labor Day (first Monday in September)
labor <- as.Date(sapply(d$yr, function(y) {
  x <- as.Date(paste0(y, "-09-01"))
  x + (8 - as.POSIXlt(x)$wday) %% 7
}))
d$labor <- ifelse(d$date == labor,1,0)

# MOTHER'S Day (2ND SUNDAY in MAY)
mother <- as.Date(sapply(d$yr, function(y) {
  x <- as.Date(paste0(y, "-05-01"))
  x + (8 - as.POSIXlt(x)$wday) %% 7 + 6
}))
d$mother <- ifelse(d$date == mother,1,0)

# FATHER'S Day (3RD SUNDAY in JUNE)
father <- as.Date(sapply(d$yr, function(y) {
  x <- as.Date(paste0(y, "-06-01"))
  x + (8 - as.POSIXlt(x)$wday) %% 7 + 13
}))
d$father <- ifelse(d$date == father,1,0)

# single holiday variable (indicator of any holiday) ADDED JUNETEENTH
d$holiday <- as.integer(d$july4 | d$memorial | d$labor | d$juneteenth | d$mother | d$father)
```


QP adjusting for Holiday Confounders (Lumped)
```{r}
# Quasi-Poisson regression: all holidays combined
quasi_holiday_model <- glm(
  HEAT_ED_VISIT_COUNT ~ MAX_DAILY_TEMP + holiday,
  family = quasipoisson(link = "log"),
  data = d
)

# Graph
plot(
  d$MAX_DAILY_TEMP,
  d$HEAT_ED_VISIT_COUNT,
  pch = 16,
  col = "black",
  xlab = "Maximum Daily Temperature",
  ylab = "Heat ED Visit Count",
  main = "Quasi-Poisson Adjusted for Holidays"
)

# Predicted curve (non-holiday days)
temp_seq <- seq(
  min(d$MAX_DAILY_TEMP),
  max(d$MAX_DAILY_TEMP),
  length.out = 100
)

predicted <- predict(
  quasi_holiday_model,
  newdata = data.frame(
    MAX_DAILY_TEMP = temp_seq,
    holiday = 0
  ),
  type = "response"
)

lines(
  temp_seq,
  predicted,
  col = "red",
  lwd = 2
)
```

QP + Holiday Confounders Summary (Lumped)
```{r}
summary(quasi_holiday_model)
```

QP adjusting for Holiday Confounders (Seperate)
```{r}
# Quasi-Poisson regression: individual holidays
quasi_individual_holiday_model <- glm(
  HEAT_ED_VISIT_COUNT ~ MAX_DAILY_TEMP +
    july4 +
    juneteenth +
    memorial +
    labor +
    mother +
    father,
  family = quasipoisson(link = "log"),
  data = d
)

# Graph
plot(
  d$MAX_DAILY_TEMP,
  d$HEAT_ED_VISIT_COUNT,
  pch = 16,
  col = "black",
  xlab = "Maximum Daily Temperature",
  ylab = "Heat ED Visit Count",
  main = "Quasi-Poisson Adjusted for Individual Holidays"
)

# Predicted curve (typical non-holiday day)
temp_seq <- seq(
  min(d$MAX_DAILY_TEMP),
  max(d$MAX_DAILY_TEMP),
  length.out = 100
)

predicted <- predict(
  quasi_individual_holiday_model,
  newdata = data.frame(
    MAX_DAILY_TEMP = temp_seq,
    july4 = 0,
    juneteenth = 0,
    memorial = 0,
    labor = 0,
    mother = 0,
    father = 0
  ),
  type = "response"
)

lines(
  temp_seq,
  predicted,
  col = "red",
  lwd = 2
)
```

QP + Holiday Confounders Summary (Seperate)
```{r}
summary(quasi_individual_holiday_model)
```

Comparing Lumped vs Seperate 
```{r}
# ==========================================
# FIT STATISTICS:
# Lumped holiday vs Individual holidays
# ==========================================

holiday_fit_compare <- data.frame(

  Model = c(
    "Lumped Holiday",
    "Individual Holidays"
  ),

  Residual_Deviance = c(
    deviance(quasi_holiday_model),
    deviance(quasi_individual_holiday_model)
  ),

  Residual_DF = c(
    df.residual(quasi_holiday_model),
    df.residual(quasi_individual_holiday_model)
  ),

  Dispersion = c(
    summary(quasi_holiday_model)$dispersion,
    summary(quasi_individual_holiday_model)$dispersion
  )
)


# View fit statistics side-by-side

holiday_fit_compare


# Test whether individual holidays improve fit

anova(
  quasi_holiday_model,
  quasi_individual_holiday_model,
  test = "F"
)

#Separating holidays doesn't make a big difference 
```



Example Code - LAG CODE
```{r}
# Create daily dates for June, 2019-2023 - CHANGE (2017-2025)
date <- seq(as.Date("2017-05-01"), as.Date("2025-09-30"), by = "day")
date <- date[format(date, "%m") %in% c("05","06","07","08","09")]


#Deleted Simulated Temp & Data Frame 

# Split temperatures by year so lags reset each May-September
d$yr <- format(d$date, "%Y")
temp_by_year <- split(d$MAX_DAILY_TEMP, yr)



#Quasi-Poisson regression
quasi_pois_model_yr <- glm(
  HEAT_ED_VISIT_COUNT ~ MAX_DAILY_TEMP+as.integer(yr),
  family = quasipoisson(link = "log"),
  data = d
)
summary(quasi_pois_model_yr)


quasi_pois_model_yrf <- glm(
  HEAT_ED_VISIT_COUNT ~ MAX_DAILY_TEMP+as.factor(yr),
  family = quasipoisson(link = "log"),
  data = d
)
summary(quasi_pois_model_yrf)

summary(quasi_pois_model)

#Graphing 
plot(
  d$MAX_DAILY_TEMP,
  d$HEAT_ED_VISIT_COUNT,
  pch = 16,
  col = "black",
  xlab = "Maximum Daily Temperature",
  ylab = "Heat ED Visit Count",
  main = "Quasi-Poisson Regression"
)

#Line
lines(
  sort(d$MAX_DAILY_TEMP),
  fitted(quasi_pois_model)[order(d$MAX_DAILY_TEMP)],
  col = "red",
  lwd = 2
)




# Create lag0 through lag6 within each year
for (i in 0:6) {
  d[[paste0("lag", i)]] <- unlist(
    lapply(temp_by_year, function(x) {
      if (i == 0) {
        x
      } else {
        c(rep(NA, i), head(x, -i))
      }
    }),
    use.names = FALSE
  )
}

head(d[, c("date", "MAX_DAILY_TEMP",
           "lag1",
           "lag6")])

```

QP adjusting for Holiday (Lumped+Lags)
```{r}
# Quasi-Poisson regression:
# Adjusted for same-day temperature, lag1, lag6, and lumped holiday

quasi_lag_holiday_model <- glm(
  HEAT_ED_VISIT_COUNT ~ MAX_DAILY_TEMP +
    lag1 +
    lag6 +
    holiday,
  family = quasipoisson(link = "log"),
  data = d
)

# Plot observed data
plot(
  d$MAX_DAILY_TEMP,
  d$HEAT_ED_VISIT_COUNT,
  pch = 16,
  col = "black",
  xlab = "Maximum Daily Temperature",
  ylab = "Heat ED Visit Count",
  main = "Quasi-Poisson: Lags + Holiday Adjustment"
)


# Temperature range

temp_seq <- seq(
  min(d$MAX_DAILY_TEMP, na.rm = TRUE),
  max(d$MAX_DAILY_TEMP, na.rm = TRUE),
  length.out = 100
)


# Prediction:
# Average lag temperatures + average holiday frequency

predicted_adjusted <- predict(
  quasi_lag_holiday_model,
  newdata = data.frame(
    MAX_DAILY_TEMP = temp_seq,
    lag1 = mean(d$lag1, na.rm = TRUE),
    lag6 = mean(d$lag6, na.rm = TRUE),
    holiday = mean(d$holiday, na.rm = TRUE)
  ),
  type = "response"
)


# Add adjusted regression line

lines(
  temp_seq,
  predicted_adjusted,
  col = "red",
  lwd = 2
)
```

QP adjusting for Holiday AVG (Lumped+Lags)  
```{r}
# ==========================================
# Plot observed data
# ==========================================

plot(
  d$MAX_DAILY_TEMP,
  d$HEAT_ED_VISIT_COUNT,
  pch = 16,
  col = "black",
  xlab = "Maximum Daily Temperature",
  ylab = "Heat ED Visit Count",
  main = "Quasi-Poisson: Lags + Holiday Adjustment"
)


# ==========================================
# Create prediction line
# ==========================================

temp_seq <- seq(
  min(d$MAX_DAILY_TEMP, na.rm = TRUE),
  max(d$MAX_DAILY_TEMP, na.rm = TRUE),
  length.out = 100
)


predicted_average <- predict(
  quasi_lag_holiday_model,
  newdata = data.frame(
    MAX_DAILY_TEMP = temp_seq,
    lag1 = mean(d$lag1, na.rm = TRUE),
    lag6 = mean(d$lag6, na.rm = TRUE),
    holiday = mean(d$holiday, na.rm = TRUE)
  ),
  type = "response"
)


# ==========================================
# Add red regression line
# ==========================================

lines(
  temp_seq,
  predicted_average,
  col = "red",
  lwd = 2
)
```

QP adjusting for Holiday (Lumped+Lags) Summary 
```{r}
summary(quasi_lag_holiday_model)
```


Compare Holiday vs Holiday w/Lag
```{r}
# ==========================================
# Compare holiday-adjusted vs lag-adjusted regression lines
# ==========================================

# Temperature sequence

temp_seq <- seq(
  min(d$MAX_DAILY_TEMP, na.rm = TRUE),
  max(d$MAX_DAILY_TEMP, na.rm = TRUE),
  length.out = 100
)


# Prediction: temperature + holiday

pred_holiday <- predict(
  quasi_holiday_model,
  newdata = data.frame(
    MAX_DAILY_TEMP = temp_seq,
    holiday = mean(d$holiday, na.rm = TRUE)
  ),
  type = "response"
)


# Prediction: temperature + holiday + lags

pred_lag_holiday <- predict(
  quasi_lag_holiday_model,
  newdata = data.frame(
    MAX_DAILY_TEMP = temp_seq,
    lag1 = mean(d$lag1, na.rm = TRUE),
    lag6 = mean(d$lag6, na.rm = TRUE),
    holiday = mean(d$holiday, na.rm = TRUE)
  ),
  type = "response"
)


# Plot observed data

plot(
  d$MAX_DAILY_TEMP,
  d$HEAT_ED_VISIT_COUNT,
  pch = 16,
  col = "gray",
  xlab = "Maximum Daily Temperature",
  ylab = "Heat ED Visit Count",
  main = "Effect of Lag Adjustment on Quasi-Poisson Model"
)


# Add regression lines

lines(
  temp_seq,
  pred_holiday,
  col = "red",
  lwd = 2
)

lines(
  temp_seq,
  pred_lag_holiday,
  col = "blue",
  lwd = 2
)


# Legend

legend(
  "topleft",
  legend = c(
    "Temperature + Holiday",
    "Temperature + Holiday + Lag1 + Lag6"
  ),
  col = c("red", "blue"),
  lwd = 2,
  cex = 0.8
)
```




Advanced options:
Consider nonlinear exposure-response relations, for example, through the use of polynomial terms or splines


Example Code - Nonlinear effect of temperature - SAVE FOR LATER 
```{r}
# ==========================================
# Negative binomial spline models
# Using your variables:
# Outcome = HEAT_ED_VISIT_COUNT
# Exposure = MAX_DAILY_TEMP
# ==========================================

library(MASS)
library(splines)


# ------------------------------------------
# Linear temperature effect
# ------------------------------------------

fit1 <- glm.nb(
  HEAT_ED_VISIT_COUNT ~ MAX_DAILY_TEMP,
  data = d
)


# ------------------------------------------
# Natural spline with 3 df
# Knots chosen automatically
# ------------------------------------------

fit2 <- glm.nb(
  HEAT_ED_VISIT_COUNT ~ ns(MAX_DAILY_TEMP, df = 3),
  data = d
)


# ------------------------------------------
# Natural spline with user-specified knots
# 50th and 85th percentile
# ------------------------------------------

fit3 <- glm.nb(
  HEAT_ED_VISIT_COUNT ~ ns(
    MAX_DAILY_TEMP,
    knots = quantile(
      MAX_DAILY_TEMP,
      c(0.5, 0.85),
      na.rm = TRUE
    )
  ),
  data = d
)


# ------------------------------------------
# Natural spline with 4 df
# ------------------------------------------

fit4 <- glm.nb(
  HEAT_ED_VISIT_COUNT ~ ns(MAX_DAILY_TEMP, df = 4),
  data = d
)


# ==========================================
# Compare model fit using AIC
# Lower AIC = better fit
# ==========================================

AIC(fit1, fit2, fit3, fit4)



# ==========================================
# Likelihood ratio tests
# Compare spline models to linear model
# ==========================================

anova(fit1, fit2, test = "Chisq")

anova(fit1, fit3, test = "Chisq")

anova(fit1, fit4, test = "Chisq")



# ==========================================
# Inspect spline knot locations
# ==========================================

attr(
  ns(d$MAX_DAILY_TEMP, df = 3),
  "knots"
)


attr(
  ns(d$MAX_DAILY_TEMP, df = 4),
  "knots"
)


attr(
  ns(
    d$MAX_DAILY_TEMP,
    knots = quantile(
      d$MAX_DAILY_TEMP,
      c(0.5, 0.85),
      na.rm = TRUE
    )
  ),
  "knots"
)



# ==========================================
# Create temperature grid for prediction
# ==========================================

temp_grid <- seq(
  min(d$MAX_DAILY_TEMP, na.rm = TRUE),
  max(d$MAX_DAILY_TEMP, na.rm = TRUE),
  length.out = 200
)


nd <- data.frame(
  MAX_DAILY_TEMP = temp_grid
)



# ==========================================
# Plot 1: Response scale
# Expected ED visits
# ==========================================

p1 <- predict(
  fit1,
  newdata = nd,
  type = "response"
)

p2 <- predict(
  fit2,
  newdata = nd,
  type = "response"
)

p3 <- predict(
  fit3,
  newdata = nd,
  type = "response"
)

p4 <- predict(
  fit4,
  newdata = nd,
  type = "response"
)


plot(
  d$MAX_DAILY_TEMP,
  d$HEAT_ED_VISIT_COUNT,
  pch = 16,
  col = "grey70",
  xlab = "Maximum Daily Temperature",
  ylab = "Daily Heat ED Visits",
  main = "Negative Binomial Spline Models"
)


lines(temp_grid, p1, col = "black", lwd = 2)

lines(temp_grid, p2, col = "blue", lwd = 2)

lines(temp_grid, p3, col = "red", lwd = 2, lty = 2)

lines(temp_grid, p4, col = "darkgreen", lwd = 2, lty = 3)


legend(
  "topleft",
  legend = c(
    "Linear",
    "ns(df = 3)",
    "ns(knots = 0.5, 0.85)",
    "ns(df = 4)"
  ),
  col = c(
    "black",
    "blue",
    "red",
    "darkgreen"
  ),
  lwd = 2,
  lty = c(1,1,2,3),
  bty = "n"
)



# ==========================================
# Plot 2: Link scale
# Log expected ED visits
# ==========================================

p1 <- predict(
  fit1,
  newdata = nd,
  type = "link"
)

p2 <- predict(
  fit2,
  newdata = nd,
  type = "link"
)

p3 <- predict(
  fit3,
  newdata = nd,
  type = "link"
)

p4 <- predict(
  fit4,
  newdata = nd,
  type = "link"
)


plot(
  d$MAX_DAILY_TEMP,
  log(d$HEAT_ED_VISIT_COUNT + 0.8),
  pch = 16,
  col = "grey70",
  ylim = c(-0.8,5),
  xlab = "Maximum Daily Temperature",
  ylab = "Log(Daily Heat ED Visits + 0.8)",
  main = "Negative Binomial Spline Models: Link Scale"
)


lines(temp_grid, p1, col = "black", lwd = 2)

lines(temp_grid, p2, col = "blue", lwd = 2)

lines(temp_grid, p3, col = "red", lwd = 2, lty = 2)

lines(temp_grid, p4, col = "darkgreen", lwd = 2, lty = 3)


legend(
  "topleft",
  legend = c(
    "Linear",
    "ns(df = 3)",
    "ns(knots = 0.5, 0.85)",
    "ns(df = 4)"
  ),
  col = c(
    "black",
    "blue",
    "red",
    "darkgreen"
  ),
  lwd = 2,
  lty = c(1,1,2,3),
  bty = "n"
)
```


Consider lagged effects, since heat exposure may affect health immediately or after several days
