---
title: "Analysis"
author: "LZM"
date: "2026-06-30"
output: html_document
---

Loading  
```{r}
#install.packages('tidyverse')
  library(tidyverse)
#install.packages('dplyr')
  library(dplyr)
#install.packages('gmodels')
  library(gmodels)
#install.packages('readxl')
  library(readxl)
#install.packages('lubridate')
  library(lubridate)
#install.packages('tableone')
  library(tableone)
#install.packages('knitr')
  library(knitr)
```

WD
```{r}
getwd()
```

Merging Data - DOESN'T WORK WITH CONFOUNDER CODE - 1
```{r}

data <- read.csv("/Users/lilianamorales/Heat Project/previous_years.csv")
data1 <- read.csv("/Users/lilianamorales/Heat Project/edheat2025_live.csv")


data2 <- read.csv("previous_years.csv")

data1 <- read.csv("edheat2025_live.csv")

data <- bind_rows(data2, data1)
#Combines data sets with similar rows and columns

data$Date <- as.Date(data$END_DATE)
#Turns end date column into date

data$END_DATE <- NULL
```


Merge Data - WORKS WITH PROVIDED CON CODE - 2
```{r}
library(dplyr)

# Read the data
data2 <- read.csv("previous_years.csv", stringsAsFactors = FALSE)
data1 <- read.csv("edheat2025_live.csv", stringsAsFactors = FALSE)

# Combine the datasets
d <- bind_rows(data2, data1)

# Convert END_DATE to a Date object and name it 'date'
d$date <- as.Date(d$END_DATE)

# Remove the original column if you no longer need it
d$END_DATE <- NULL
```



Step 4: Crude Analyses
Correlation Analysis
To evaluate the crude (unadjusted) relation between temperature and heat-related ED visits, calculate:
Pearson correlation|
```{r}
cor.test(
  data$MAX_DAILY_TEMP,
  data$HEAT_ED_VISIT_COUNT,
  method = "pearson"
)
```


Spearman correlation

```{r}
cor.test(
  data$MAX_DAILY_TEMP,
  data$HEAT_ED_VISIT_COUNT,
  method = "spearman"
)
```



Simple Regression
Fit a crude model: Heat-related ED visits = f(temperature)
You might explore different possible approaches, Linear regression
```{r}
model1 <- lm(HEAT_ED_VISIT_COUNT ~ MAX_DAILY_TEMP, data = data)

summary(model1)
```


Poisson regression
```{r}
glm(
  HEAT_ED_VISIT_COUNT ~ MAX_DAILY_TEMP,
  family = poisson(link = "log"),
  data = data
)
```
Quasi Poisson regression
```{r}
glm(
  HEAT_ED_VISIT_COUNT ~ MAX_DAILY_TEMP,
  family = quasipoisson(link = "log"),
  data = data
)
```



Negative binomial regression
First Load MASS 
```{r}
library(MASS)
```

Negative binomial regression
```{r}
glm.nb(
  HEAT_ED_VISIT_COUNT ~ MAX_DAILY_TEMP,
  data = data
)
```

Discuss advantages and limitations of each approach.








Step 5: Time-Series Analysis
Build and refine your analytical models. Consider whether adjustment for time-varying factors (e.g., potential confounders), seasonality, temporal trends, or other covariates is warranted. To adjust for time, consider adding variables such as year, day of season, long-term trend. Clearly justify your modeling choices and discuss any assumptions made.


Extreme Heat Days - WORK ON IT & FIX
```{r}
```




Identified Holidays 
- Independence Day - JULY 4
- Juneteenth - JUNE 19
- Memorial Day - LAST MON. MAY 
- Labor Day - 1ST MON. SEP
- Mother's Day - 2ND SUN. MAY 
- Father's Day - 3RD SUN. JUNE


Example Code 2 - Holidays 
```{r}
# ensure date is in Date format
d$date <- as.Date(d$date)

# create a couple calculated date variables
# year
d$yr <- as.integer(format(d$date, "%Y"))

# month
d$mo <- as.integer(format(d$date, "%m"))

# day of week
d$dow <- weekdays(d$date)
d$dow <- factor(d$dow, levels=unique(d$dow))

# day of the warm season (integer, restarts at 1 each year)
d$season_day <- as.integer(d$date - as.Date(paste0(format(d$date, "%Y"), "-05-01"))) + 1

# create indicator for three summer holidays
# July 4
d$july4 <- ifelse(format(d$date, "%m-%d") == "07-04",1,0)

#ADDING JUNETEENTH
d$juneteenth <- ifelse(format(d$date, "%m-%d") == "06-19",1,0)

# Memorial Day (last Monday in May)
memorial <- as.Date(sapply(d$yr, function(y) {
  x <- as.Date(paste0(y, "-05-31"))
  x - as.POSIXlt(x)$wday   # subtract 0-6 days to reach Monday
}))
d$memorial <- ifelse(d$date == memorial,1,0)

# Labor Day (first Monday in September)
labor <- as.Date(sapply(d$yr, function(y) {
  x <- as.Date(paste0(y, "-09-01"))
  x + (8 - as.POSIXlt(x)$wday) %% 7
}))
d$labor <- ifelse(d$date == labor,1,0)

# MOTHER'S Day (2ND SUNDAY in MAY)
mother <- as.Date(sapply(d$yr, function(y) {
  x <- as.Date(paste0(y, "-05-01"))
  x + (8 - as.POSIXlt(x)$wday) %% 7 + 7
}))
d$mother <- ifelse(d$date == mother,1,0)

# FATHER'S Day (3RD SUNDAY in JUNE)
father <- as.Date(sapply(d$yr, function(y) {
  x <- as.Date(paste0(y, "-06-01"))
  x + (8 - as.POSIXlt(x)$wday) %% 7 + 14
}))
d$father <- ifelse(d$date == father,1,0)

# single holiday variable (indicator of any holiday) ADDED JUNETEENTH
d$holiday <- as.integer(d$july4 | d$memorial | d$labor | d$juneteenth | d$mother | d$father)

```



Example Code - LAG CODE
```{r}
set.seed(123)

# Create daily dates for June, 2019-2023 - DONE FOR MONTHS MAY-SEP. 2017-2025
#MAY
datemay <- seq(as.Date("2017-05-01"), as.Date("2025-05-31"), by = "day")
datemay <- date[format(date, "%m") == "06"]
#JUNE
datejune <- seq(as.Date("2017-06-01"), as.Date("2025-06-30"), by = "day")
datejune <- date[format(date, "%m") == "06"]
#JULY 
datejuly <- seq(as.Date("2017-07-01"), as.Date("2025-07-31"), by = "day")
datejuly <- date[format(date, "%m") == "06"]
#AUGUST 
dateaugust <- seq(as.Date("2017-08-01"), as.Date("2025-08-31"), by = "day")
dateaugust <- date[format(date, "%m") == "06"]
#SEPTEMBER 
dateseptember <- seq(as.Date("2017-09-01"), as.Date("2025-09-30"), by = "day")
dateseptember <- date[format(date, "%m") == "06"]


# Simulate temperatures
temperature <- round(rnorm(length(date), mean = 75, sd = 8), 1)

# Create data frame
dat <- data.frame(date = date, temperature = temperature)

# Split temperatures by year so lags reset each June
yr <- format(dat$date, "%Y")
temp_by_year <- split(dat$temperature, yr)

# Create lag0 through lag6 within each year
for (i in 0:6) {
  dat[[paste0("lag", i)]] <- unlist(
    lapply(temp_by_year, function(x) {
      if (i == 0) {
        x
      } else {
        c(rep(NA, i), head(x, -i))
      }
    }),
    use.names = FALSE
  )
}

head(dat, 45)
```


Example Code - Nonlinear effect of temperature - SAVE FOR LATER
```{r}
# fit models
library(MASS)
library(splines)

# negative binomial models

# linear temperature effects
fit1 <- glm.nb(visits ~ temp, data = d)

# Natural spline with 3 df; 2 knots chosen automatically
fit2 <- glm.nb(visits ~ ns(temp, df = 3), data = d)

# Natural spline with 2 user-specified knot locations
fit3 <- glm.nb(
  visits ~ ns(temp, knots = quantile(temp, c(0.5, 0.85), na.rm = TRUE)),
  data = d
)

# Natural spline with 4 df; 3 knots chosen automatically
fit4 <- glm.nb(visits ~ ns(temp, df = 4), data = d)

# Compare overall model fit using AIC (aim to minimize)
AIC(fit1, fit2, fit3, fit4)

# Likelihood-ratio tests versus the linear model; others aren't nested
anova(fit1, fit2, test = "Chisq")
anova(fit1, fit3, test = "Chisq")
anova(fit1, fit4, test = "Chisq")

# Inspect knot locations used by the spline bases
attr(ns(d$temp, df = 3), "knots") # 33rd and 67th percentiles (automatic)
attr(ns(d$temp, df = 4), "knots")
attr(ns(d$temp, knots = quantile(d$temp, c(0.5, 0.85), na.rm = TRUE)), "knots") # 50th and 85th percentiles (user specified)

# Grid of temperatures for smooth prediction curves
temp_grid <- seq(min(d$temp, na.rm = TRUE),
                 max(d$temp, na.rm = TRUE),
                 length.out = 200)

nd <- data.frame(temp = temp_grid)

par(mfrow = c(1, 2))

# -----------------------------
# Plot 1: response scale
# -----------------------------
p1 <- predict(fit1, newdata = nd, type = "response")
p2 <- predict(fit2, newdata = nd, type = "response")
p3 <- predict(fit3, newdata = nd, type = "response")
p4 <- predict(fit4, newdata = nd, type = "response")

plot(d$temp, d$visits,
     pch = 16, col = "grey70",
     xlab = "Temperature", ylab = "Daily ED visits",
     main = "Response scale")

lines(temp_grid, p1, col = "black", lwd = 2)
lines(temp_grid, p2, col = "blue",  lwd = 2)
lines(temp_grid, p3, col = "red",   lwd = 2, lty = 2)
lines(temp_grid, p4, col = "darkgreen", lwd = 2, lty = 3)

legend("topleft",
       legend = c("Linear", "ns(df = 3)", "ns(knots = c(0.5, 0.85))", "ns(df = 4)"),
       col = c("black", "blue", "red", "darkgreen"),
       lwd = 2, lty = c(1, 1, 2, 3), bty = "n")

# -----------------------------
# Plot 2: link scale
# -----------------------------
p1 <- predict(fit1, newdata = nd, type = "link")
p2 <- predict(fit2, newdata = nd, type = "link")
p3 <- predict(fit3, newdata = nd, type = "link")
p4 <- predict(fit4, newdata = nd, type = "link")

plot(d$temp, log(d$visits + 0.8),
     pch = 16, col = "grey70", ylim=c(-.8,5),
     xlab = "Temperature", ylab = "log(Daily ED visits + 0.8)",
     main = "Link scale")

lines(temp_grid, p1, col = "black", lwd = 2)
lines(temp_grid, p2, col = "blue",  lwd = 2)
lines(temp_grid, p3, col = "red",   lwd = 2, lty = 2)
lines(temp_grid, p4, col = "darkgreen", lwd = 2, lty = 3)

legend("topleft",
       legend = c("Linear", "ns(df = 3)", "ns(knots = c(0.5, 0.85))", "ns(df = 4)"),
       col = c("black", "blue", "red", "darkgreen"),
       lwd = 2, lty = c(1, 1, 2, 3), bty = "n")
```



Advanced options:
Consider nonlinear exposure-response relations, for example, through the use of polynomial terms or splines

Consider lagged effects, since heat exposure may affect health immediately or after several days
