---
title: "Heat Exposure and Emergency Department Visits in New York City: Association and Prediction of Heat-Related Health Outcomes (Final Code)"
format: html
editor: visual
---

## Merge data

```{r}
library(dplyr)
data2 <- read.csv("previous_years.csv")

data1 <- read.csv("edheat2025_live.csv")

d <- bind_rows(data2, data1)
#Combines data sets with similar rows and columns

d$date <- as.Date(d$END_DATE)
d$END_DATE <- NULL
#Date conversion
d$yr <- as.integer(format(d$date, "%Y"))

d$Month <- as.integer(format(d$date, "%m"))
#Month Variable

```

## Quassi-Poisson Regression Considering Year

```{r}
# Create daily dates for June, 2019-2023 - CHANGE (2017-2025)
date <- seq(as.Date("2017-05-01"), as.Date("2025-09-30"), by = "day")
date <- date[format(date, "%m") %in% c("05","06","07","08","09")]
yr = d$yr

#Deleted Simulated Temp & Data Frame 

# Split temperatures by year so lags reset each May-September
d$yr <- format(d$date, "%Y")
temp_by_year <- split(d$MAX_DAILY_TEMP, yr)



#Quasi-Poisson regression
quasi_pois_model_yr <- glm(
  HEAT_ED_VISIT_COUNT ~ MAX_DAILY_TEMP+yr,
  family = quasipoisson(link = "log"),
  data = d
)
summary(quasi_pois_model_yr)


quasi_pois_model_yrf <- glm(
  HEAT_ED_VISIT_COUNT ~ MAX_DAILY_TEMP+as.factor(yr),
  family = quasipoisson(link = "log"),
  data = d
)
summary(quasi_pois_model_yrf)

#summary(quasi_pois_model)

#Graphing 
plot(
  d$MAX_DAILY_TEMP,
  d$HEAT_ED_VISIT_COUNT,
  pch = 16,
  col = "black",
  xlab = "Maximum Daily Temperature",
  ylab = "Heat ED Visit Count",
  main = "Quasi-Poisson Regression Considering Year"
)

#Line
lines(
  sort(d$MAX_DAILY_TEMP),
  fitted(quasi_pois_model_yrf)[order(d$MAX_DAILY_TEMP)],
  col = "red",
  lwd = 2
)
```

## Quassi-Poisson Regression Considering Month

```{r}
d$Month <- format(d$date, "%m")
temp_by_month <- split(d$MAX_DAILY_TEMP, d$Month)


quasi_pois_model_mf <- glm(
  HEAT_ED_VISIT_COUNT ~ MAX_DAILY_TEMP+as.factor(Month),
  family = quasipoisson(link = "log"),
  data = d
)
summary(quasi_pois_model_mf)

#summary(quasi_pois_model)

#Graphing 
plot(
  d$MAX_DAILY_TEMP,
  d$HEAT_ED_VISIT_COUNT,
  pch = 16,
  col = "black",
  xlab = "Maximum Daily Temperature",
  ylab = "Heat ED Visit Count",
  main = "Quasi-Poisson Regression Considering Month"
)

#Line
lines(
  sort(d$MAX_DAILY_TEMP),
  fitted(quasi_pois_model_mf)[order(d$MAX_DAILY_TEMP)],
  col = "red",
  lwd = 2
)
```

## Creating Holiday Variables

```{r}
# Independence Day
d$july4 <- ifelse(format(d$date, "%m-%d") == "07-04",1,0)

# Juneteenth
d$juneteenth <- ifelse(format(d$date, "%m-%d") == "06-19",1,0)

# Mother's Day
mother <- as.Date(sapply(d$yr, function(y) {
  x <- as.Date(paste0(y, "-05-01"))
  x + (8 - as.POSIXlt(x)$wday) %% 7 + 6
}))
d$mother <- ifelse(d$date == mother,1,0)

# Father's Day
father <- as.Date(sapply(d$yr, function(y) {
  x <- as.Date(paste0(y, "-06-01"))
  x + (8 - as.POSIXlt(x)$wday) %% 7 + 13
}))
d$father <- ifelse(d$date == father,1,0)

# Memorial Day
memorial <- as.Date(sapply(d$yr, function(y) {
  x <- as.Date(paste0(y, "-05-01"))
  x - as.POSIXlt(x)$wday %% 7 + 29
}))
d$memorial <- ifelse(d$date == memorial,1,0)

# Labor Day
labor <- as.Date(sapply(d$yr, function(y) {
  x <- as.Date(paste0(y, "-09-01"))
  x + (8 - as.POSIXlt(x)$wday) %% 7
}))
d$labor <- ifelse(d$date == labor,1,0)

# single holiday
d$holiday <- as.integer(d$july4 | d$memorial | d$labor | d$juneteenth | d$mother | d$father)
```

## Creating Lags 0-6

```{r}
# Create lag0 through lag6 within each year
for (i in 0:6) {
  d[[paste0("lag", i)]] <- unlist(
    lapply(temp_by_year, function(x) {
      if (i == 0) {
        x
      } else {
        c(rep(NA, i), head(x, -i))
      }
    }),
    use.names = FALSE
  )
}

head(d[, c("date", "MAX_DAILY_TEMP",
           "lag1",
           "lag6")])
```

## Quassi-Poisson Regression Using All Covariates + Lag 0

```{r}
#install.packages('car')
library(car)
# vif(quasi_pois_model_all)
quasi_pois_model_all <- glm(
  HEAT_ED_VISIT_COUNT ~ MAX_DAILY_TEMP+as.factor(d$Month)+d$holiday+as.factor(yr),
  family = quasipoisson(link = "log"),
  data = d
)
summary(quasi_pois_model_all)

#Graphing 
plot(
  d$MAX_DAILY_TEMP,
  d$HEAT_ED_VISIT_COUNT,
  pch = 16,
  col = "black",
  xlab = "Maximum Daily Temperature",
  ylab = "Heat ED Visit Count",
  main = "Quasi-Poisson Regression with Covariates"
)

#Line
lines(
  sort(d$MAX_DAILY_TEMP),
  fitted(quasi_pois_model_all)[order(d$MAX_DAILY_TEMP)],
  col = "red",
  lwd = 2
)
```

## Quassi-Poisson Regression Considering Lag 1-6 independently

```{r}
# Lag 1
quasi_pois_model_lag1 <- glm(
  HEAT_ED_VISIT_COUNT ~ lag1+as.factor(Month)+holiday+as.factor(yr),
  family = quasipoisson(link = "log"),
  data = d
)
summary(quasi_pois_model_lag1)

# Lag 2
quasi_pois_model_lag2 <- glm(
  HEAT_ED_VISIT_COUNT ~ lag2+as.factor(Month)+holiday+as.factor(yr),
  family = quasipoisson(link = "log"),
  data = d
)
summary(quasi_pois_model_lag2)

# Lag 3
quasi_pois_model_lag3 <- glm(
  HEAT_ED_VISIT_COUNT ~ lag3+as.factor(Month)+holiday+as.factor(yr),
  family = quasipoisson(link = "log"),
  data = d
)
summary(quasi_pois_model_lag3)

# Lag 4
quasi_pois_model_lag4 <- glm(
  HEAT_ED_VISIT_COUNT ~ lag4+as.factor(Month)+holiday+as.factor(yr),
  family = quasipoisson(link = "log"),
  data = d
)
summary(quasi_pois_model_lag4)

# Lag 5
quasi_pois_model_lag5 <- glm(
  HEAT_ED_VISIT_COUNT ~ lag5+as.factor(Month)+holiday+as.factor(yr),
  family = quasipoisson(link = "log"),
  data = d
)
summary(quasi_pois_model_lag5)

# Lag 6
quasi_pois_model_lag6 <- glm(
  HEAT_ED_VISIT_COUNT ~ lag6+as.factor(Month)+holiday+as.factor(yr),
  family = quasipoisson(link = "log"),
  data = d
)
summary(quasi_pois_model_lag6)
```

## Quassi-Poisson Regression Using All Covariates + Lag 0 + Natural Spline

```{r}
library(splines)

# Fit the model
quasi_spline <- glm(
  HEAT_ED_VISIT_COUNT ~ ns(MAX_DAILY_TEMP, df = 4) +
    as.factor(Month) +
    holiday +
    as.factor(yr),
  family = quasipoisson(link = "log"),
  data = d
)

# Create prediction grid
plot_data <- data.frame(
  MAX_DAILY_TEMP = seq(
    min(d$MAX_DAILY_TEMP),
    max(d$MAX_DAILY_TEMP),
    length.out = 200
  ),
  Month = "07",      # July
  holiday = 0,    # Non-holiday
  yr = 2023       # Pick one year from your data
)

# Predict expected ED visits
plot_data$pred <- predict(
  quasi_spline,
  newdata = plot_data,
  type = "response"
)


# Scatterplot
plot(
  d$MAX_DAILY_TEMP,
  d$HEAT_ED_VISIT_COUNT,
  pch = 16,
  col = "black",
  xlab = "Maximum Daily Temperature (lag 0)",
  ylab = "Heat ED Visit Count",
  main = "Quasi-Poisson Regression with Natural Spline"
)

# Add smooth spline curve
lines(
  plot_data$MAX_DAILY_TEMP,
  plot_data$pred,
  col = "red",
  lwd = 3
)


anova(quasi_spline, quasi_pois_model_all, test = "F")
summary(quasi_spline)
```

## Confidence Intervals of Quassi-Poisson Regression Using All Covariates + Lag 0 + Natural Spline

```{r}
ci_log <- confint.default(quasi_pois_model_all)
print(ci_log)

ci_irr <- ((exp(ci_log))*100)
print(ci_irr)
```

| Coefficient            | Estimate | Pr(\|t\|) | 95% CI         |
|------------------------|----------|-----------|----------------|
| MAX_DAILY_TEMP (lag 0) | 15.7%    | \<.001    | (15.0%, 16.4%) |

## D3 Data (2026 new data)

```{r}
d3 <- read.csv("edheat2026_live.csv")

#Date, Month, And Year Variable

d3$date <- as.Date(d3$END_DATE)
d3$END_DATE <- NULL

d3$yr <- as.numeric(format(d3$date, "%Y"))

d3$Month <- as.numeric(format(d3$date, "%m"))

#Holiday

# Independence Day
d3$july4 <- ifelse(format(d3$date, "%m-%d") == "07-04",1,0)

# Juneteenth
d3$juneteenth <- ifelse(format(d3$date, "%m-%d") == "06-19",1,0)

# Mother's Day
mother <- as.Date(sapply(d3$yr, function(y) {
  x <- as.Date(paste0(y, "-05-01"))
  x + (8 - as.POSIXlt(x)$wday) %% 7 + 6
}))
d3$mother <- ifelse(d3$date == mother,1,0)

# Father's Day
father <- as.Date(sapply(d3$yr, function(y) {
  x <- as.Date(paste0(y, "-06-01"))
  x + (8 - as.POSIXlt(x)$wday) %% 7 + 13
}))
d3$father <- ifelse(d3$date == father,1,0)

# Memorial Day
memorial <- as.Date(sapply(d3$yr, function(y) {
  x <- as.Date(paste0(y, "-05-01"))
  x - as.POSIXlt(x)$wday %% 7 + 29
}))
d3$memorial <- ifelse(d3$date == memorial,1,0)

# Labor Day
labor <- as.Date(sapply(d3$yr, function(y) {
  x <- as.Date(paste0(y, "-09-01"))
  x + (8 - as.POSIXlt(x)$wday) %% 7
}))
d3$labor <- ifelse(d3$date == labor,1,0)

# single holiday
d3$holiday <- as.integer(d3$july4 | d3$memorial | d3$labor | d3$juneteenth | d3$mother | d3$father)


#quasi pois model for 2026
quasi_pois_model_all3 <- glm(
  HEAT_ED_VISIT_COUNT ~ MAX_DAILY_TEMP + Month + holiday,
  family = quasipoisson(link = "log"),
  data = d3
)

summary(quasi_pois_model_all3)
```

## Prediction for D3 (2026) data (Ph1) using Lag 0 ONLY

```{r}
d$yr <- as.numeric(as.character(d$yr))
d$Month <- as.numeric(as.character(d$Month))

quasi_pois_model_all_1 <- glm(
  HEAT_ED_VISIT_COUNT ~ MAX_DAILY_TEMP + holiday + yr + Month,
  family = quasipoisson(link = "log"),
  data = d )

pred_20261 <- predict(
  quasi_pois_model_all_1,
  newdata = d3,
  type = "response"
)

d3$Ph1<- pred_20261

plot(
  d3$date,
  d3$HEAT_ED_VISIT_COUNT,
  pch = 16,
  xlab = "Date (2026)",
  ylab = "Heat-Related ED Visits",
  main = "Observed vs Predicted Heat ED Visits: 2026"
)

lines(
  d3$date,
  d3$Ph1,
  col = "red",
  lwd = 2
)

legend(
  "topleft",
  legend = c("Observed", "Predicted"),
  pch = c(16, NA),
  lty = c(NA, 1),
  col = c("black", "red")
)
```

## Prediction D3 (2026) data (Ph2) using all Lags dependently

```{r}
temp_by_year3 <- split(d3$MAX_DAILY_TEMP, d3$yr)

for (i in 0:6) {
  d3[[paste0("lag", i)]] <- unlist(
    lapply(temp_by_year3, function(x) {
      if (i == 0) {
        x
      } else {
        c(rep(NA, i), head(x, -i))
      }
    }),
    use.names = FALSE
  )
}

head(d3[, c("date", "MAX_DAILY_TEMP",
            "lag1",
            "lag6")])

quasi_pois_model_all_2 <- glm(
  HEAT_ED_VISIT_COUNT ~ MAX_DAILY_TEMP + holiday + yr + Month + lag1 + lag2 + lag3 + lag4 + lag5 + lag6,
  family = quasipoisson(link = "log"),
  data = d )

pred_20262 <- predict(
  quasi_pois_model_all_2,
  newdata = d3,
  type = "response"
)

d3$Ph2 <- pred_20262

plot(
  d3$date,
  d3$HEAT_ED_VISIT_COUNT,
  pch = 16,
  xlab = "Date (2026)",
  ylab = "Heat-Related ED Visits",
  main = "Observed vs Predicted Heat ED Visits: 2026"
)

lines(
  d3$date,
  d3$Ph2,
  col = "red",
  lwd = 2
)

legend(
  "topleft",
  legend = c("Observed", "Predicted"),
  pch = c(16, NA),
  lty = c(NA, 1),
  col = c("black", "red")
)

```

## RMSE Of Both 

```{r}
rmse_base <- sqrt(mean((d3$HEAT_ED_VISIT_COUNT-d3$Ph2)^2, na.rm = TRUE))
print("RMSE of model using all lags as covariates:")
print(rmse_base)

rmse_base1 <- sqrt(mean((d3$HEAT_ED_VISIT_COUNT-d3$Ph1)^2, na.rm = TRUE))
print("RMSE of model using lag0 as time covariate:")
print(rmse_base1)
```

| RMSE | R\^2 |
|------|------|
| 9.9  | 0.85 |

## Proctor Code (Estimation Analysis)

```{r}
# read in datasets
data2 <- read.csv("previous_years.csv")
data1 <- read.csv("edheat2025_live.csv")

#Combines data sets with similar rows and columns
d <- rbind(data2, data1)

# rename and create variables
d$temp <- d$MAX_DAILY_TEMP; d$MAX_DAILY_TEMP <- NULL
d$edvisit <- d$HEAT_ED_VISIT_COUNT; d$HEAT_ED_VISIT_COUNT <- NULL
d$date <- as.Date(d$END_DATE)
d$END_DATE <- NULL

# year and month as character; will be treated as factor
d$yr <- format(d$date, "%Y")
d$mo <- format(d$date, "%m")
#year and month, as integers
d$yr_i <- as.integer(d$yr)
d$mo_i <- as.integer(d$mo)
# day of week
d$dow <- weekdays(d$date)
d$dow <- factor(d$dow, levels=unique(d$dow))

# Create daily dates for June, 2019-2023 - CHANGE (2017-2025)
date <- seq(as.Date("2017-05-01"), as.Date("2025-09-30"), by = "day")
date <- date[format(date, "%m") %in% c("05","06","07","08","09")]


# Split temperatures by year so lags reset each May-September
temp_by_year <- split(d$temp, d$yr)

### Holidays
# Independence Day
d$july4 <- ifelse(format(d$date, "%m-%d") == "07-04",1,0)

# Juneteenth
d$juneteenth <- ifelse(format(d$date, "%m-%d") == "06-19",1,0)

# Mother's Day
mother <- as.Date(sapply(d$yr, function(y) {
  x <- as.Date(paste0(y, "-05-01"))
  x + (8 - as.POSIXlt(x)$wday) %% 7 + 6
}))
d$mother <- ifelse(d$date == mother,1,0)

# Father's Day
father <- as.Date(sapply(d$yr, function(y) {
  x <- as.Date(paste0(y, "-06-01"))
  x + (8 - as.POSIXlt(x)$wday) %% 7 + 13
}))
d$father <- ifelse(d$date == father,1,0)

# Memorial Day
memorial <- as.Date(sapply(d$yr, function(y) {
  x <- as.Date(paste0(y, "-05-01"))
  x - as.POSIXlt(x)$wday %% 7 + 29
}))
d$memorial <- ifelse(d$date == memorial,1,0)

# Labor Day
labor <- as.Date(sapply(d$yr, function(y) {
  x <- as.Date(paste0(y, "-09-01"))
  x + (8 - as.POSIXlt(x)$wday) %% 7
}))
d$labor <- ifelse(d$date == labor,1,0)

# single holiday
d$holiday <- as.integer(d$july4 | d$memorial | d$labor | d$juneteenth | d$mother | d$father) 

# Create lag0 through lagNLAG within each year
NLAG <- 6
for (i in 0:NLAG) {
  d[[paste0("temp_lag", i)]] <- unlist(
    lapply(temp_by_year, function(x) {
      if (i == 0) {
        x
      } else {
        c(rep(NA, i), head(x, -i))
      }
    }),
    use.names = FALSE
  )
}

head(d[, c("date", "temp",
           paste0("temp_lag", 0:NLAG))])


# Quasi-Poisson regression relating ED visit count to temperature

# fit "crude" unadjusted models, lag 0 to NLAG (alternatively could write a for loop...)
fit_lag0_crude <- glm(edvisit ~ temp_lag0, family = quasipoisson(link = "log"),data = d)
fit_lag1_crude <- glm(edvisit ~ temp_lag1, family = quasipoisson(link = "log"),data = d)
fit_lag2_crude <- glm(edvisit ~ temp_lag2, family = quasipoisson(link = "log"),data = d)
fit_lag3_crude <- glm(edvisit ~ temp_lag3, family = quasipoisson(link = "log"),data = d)
fit_lag4_crude <- glm(edvisit ~ temp_lag4, family = quasipoisson(link = "log"),data = d)
fit_lag5_crude <- glm(edvisit ~ temp_lag5, family = quasipoisson(link = "log"),data = d)
fit_lag6_crude <- glm(edvisit ~ temp_lag6, family = quasipoisson(link = "log"),data = d)

# fit adjusted models, lag 0 to NLAG [still needs holiday indicator added]
fit_lag0_adj <- glm(edvisit ~ temp_lag0+yr+mo+holiday, family = quasipoisson(link = "log"),data = d)
fit_lag1_adj <- glm(edvisit ~ temp_lag1+yr+mo+holiday, family = quasipoisson(link = "log"),data = d)
fit_lag2_adj <- glm(edvisit ~ temp_lag2+yr+mo+holiday, family = quasipoisson(link = "log"),data = d)
fit_lag3_adj <- glm(edvisit ~ temp_lag3+yr+mo+holiday, family = quasipoisson(link = "log"),data = d)
fit_lag4_adj <- glm(edvisit ~ temp_lag4+yr+mo+holiday, family = quasipoisson(link = "log"),data = d)
fit_lag5_adj <- glm(edvisit ~ temp_lag5+yr+mo+holiday, family = quasipoisson(link = "log"),data = d)
fit_lag6_adj <- glm(edvisit ~ temp_lag6+yr+mo+holiday, family = quasipoisson(link = "log"),data = d)


# results below are direct "beta" values, you still need to transform them to be more interpretable (scale and exponentiate!)

# est and 95% CI for a single model
c(fit_lag0_crude$coef[2], confint(fit_lag0_crude)[2,])

# or for all 7 models at once, and also p-value
# here's example code of how to extract and save the results
results_crude <- matrix(NA, nrow=NLAG+1, ncol=4)
colnames(results_crude) <- c("Est","CI_lb","CI_ub","p-value")
rownames(results_crude) <- paste0("temp_lag", 0:NLAG)
results_adj <- results_crude
for(i in 0:NLAG){
  # crude model
  mymodeli <- eval(parse(text=paste0("fit_lag",i,"_crude")))
  results_crude[i+1,] <- c(mymodeli$coef[2], 
                           confint(mymodeli)[2,],
                           summary(mymodeli)$coef[2,"Pr(>|t|)"])
  # adjusted model
  mymodeli <- eval(parse(text=paste0("fit_lag",i,"_adj")))
  results_adj[i+1,] <- c(mymodeli$coef[2], 
                         confint(mymodeli)[2,],
                         summary(mymodeli)$coef[2,"Pr(>|t|)"])
}
results_crude <- exp(results_crude) - 1
results_crude <- results_crude*100
results_adj <- exp(results_adj) - 1
results_adj <- results_adj*100

print(results_adj)

##########################################
## plotting code #########################

## Lag numbers
lags <- 0:NLAG


## Put results in a list
results_crude <- data.frame(results_crude)
results_adj <- data.frame(results_adj )
res_list <- list(
  "Crude model" = results_crude,
  "Adjusted model" = results_adj
)

## Common y-axis limits
ylim <- range(
  results_crude$CI_lb, results_crude$CI_ub,
  results_adj$CI_lb, results_adj$CI_ub
)

par(mfrow = c(1, 2),
    mar = c(4, 4, 3, 1),
    oma = c(0, 0, 3, 0))


#par(mfrow = c(1,2),
#mar = c(4,4,3,1))

for(i in seq_along(res_list)) {
  
  dat <- res_list[[i]]
  
  plot(lags, dat$Est,
       type = "n",
       ylim = ylim,
       xaxt = "n",
       xlab = "Lag (days)",
       ylab = "ED % Change Per 1°F Temp. Increase",
       main = names(res_list)[i])
  
  axis(1, at = lags)
  
  ## CI polygon
  polygon(c(lags, rev(lags)),
          c(dat$CI_lb, rev(dat$CI_ub)),
          col = gray(0.85),
          border = NA)
  
  ## Estimate
  lines(lags, dat$Est, lwd = 2)
  points(lags, dat$Est, pch = 19)
  
  abline(h = 0, lty = 2, col = "gray50")
}
mtext("Estimation Model",
      side = 3,
      outer = TRUE,
      line = 1,
      cex = 1.5,
      font = 2)
```
